---
title: "get_homogeneous_m6a"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we obtain the m6a variants with m/k = 1 in a cell that occur in at least 5 cells of other cell types with an m/k ratio < 0.9.

Let's start with some preliminary tests. The distribution of the coverage at the sites with m/k = 1 is:

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(Seurat)
seurat_data <- readRDS("/GPFS/Magda_lab_permanent/maitenat/scs_yth/code/scRNA_dimplot/scDART_integrated_seurat.rds")
cell_to_cluster <- seurat_data@active.ident
cell_to_sample <- seurat_data@meta.data$orig.ident
cluster_labels <- data.frame(cell = names(cell_to_cluster), 
                              cluster = cell_to_cluster,
                             sample = cell_to_sample)

filename <- "/GPFS/Magda_lab_temp/maitenat/scs_yth/6J-3M-2ws-V2-3-YTH/bullseye/2023/6J-3M-2ws-V2-3-YTH_edit-sites_bc_filtered_all_yth_filt-all.bed"
yth_data <- read.table(filename, header = FALSE, fill = TRUE)
colnames(yth_data) <- c("chr", "start", "end", "gene", "loc", "change0", "n_yth", "dart_ratio/control_ratio", "perc", "strand", "control_freq", "control_cov", "dart_freq", "dart_cov", "change", "cell_id", "cluster", "mk_overall", "mk_mut", "mk_ratio")

immune_cells <- as.character(c(0, 1, 4, 8, 9, 10, 14, 15, 17, 18, 19, 20, 21, 24, 25, 27))
oligodendrocytes <- as.character(c(2, 3, 5, 6, 7, 12, 13, 16))
neurons <- as.character(c(11, 23))
endothelial <- "26"
astrocytes <- "22"

cluster_df <- tibble(cell_type = c(rep("immune_cells", length(immune_cells)), rep("oligodendrocytes", length(oligodendrocytes)), rep("neurons", length(neurons)), "endothelial", "astrocytes"),
                     cluster_id = c(immune_cells, oligodendrocytes, neurons, endothelial, astrocytes))

yth_data_cluster_labeled <- yth_data[!is.na(yth_data[, "cluster"]), ]
yth_data_cluster_labeled$cluster <- as.character(yth_data_cluster_labeled$cluster)

yth_data_cluster_labeled <- yth_data_cluster_labeled %>%
  left_join(cluster_df, by = c("cluster" = "cluster_id"))

summary(pull(filter(yth_data_cluster_labeled, dart_freq == 1), dart_cov))
```

For each cell type, we obtain a table with each position with m/k = 1 in at least 1 cell that is also found in at least 5 cells of another cell type with m/k<0.9.

```{r echo = FALSE, message = FALSE, warning = FALSE}
test_mk1_mk09_5cells <- function(celltype) {
  cell_thr <- 5
  
  sites <- yth_data_cluster_labeled %>%
    filter(cell_type == celltype,
           dart_freq == 1) %>%
    select(chr, start, end, gene) %>%
    distinct()
  
  common_sites_5cells <- yth_data_cluster_labeled %>%
    filter(cell_type != celltype) %>%
    semi_join(sites, by = c("chr", "start", "end")) %>%
    group_by(chr, start, end, gene) %>%
    summarise(n_cells = n_distinct(cell_id)) %>%
    filter(n_cells >= cell_thr)

  other_celltypes <- yth_data_cluster_labeled %>%
  filter(cell_type != celltype) %>%
  filter(dart_freq < 0.9)


# for sites common to sites and other_celltypes, check if there are at least 5 cells with m/k < 0.9 per cell type
sites_with_other_mk09_5cells <- sites %>%
  inner_join(other_celltypes, by = c("chr", "start", "end", "gene")) %>%
  group_by(chr, start, end, gene) %>%
  summarise(n_cells = n_distinct(cell_id)) %>%
  filter(n_cells >= cell_thr)

  print(sprintf("Number of (unique) sites with m/k = 1 in %s: %s", celltype, nrow(sites)))
  print(sprintf("Of those sites, %s are also found in at least 5 cells of another cell type", nrow(common_sites_5cells)))
  print(sprintf("Of those %s sites, %s are found with m/k < 0.9 in at least 5 cells of another cell type", nrow(common_sites_5cells), nrow(sites_with_other_mk09_5cells)))
  print(sprintf("This means that %s sites have m/k > 0.9 in at least 5 cells of another cell type", nrow(common_sites_5cells) - nrow(sites_with_other_mk09_5cells)))
  
  # save the table per cell type
  outdir <- dirname(filename)
  write.table(sites_with_other_mk09_5cells, file = file.path(outdir, paste0(celltype, "_mk1_common_mk09_5cells.txt")), sep = "\t", row.names = FALSE, quote = FALSE, col.names = TRUE)
}

# run test_mk1_mk09 for all cell types
for (celltype in unique(cluster_df$cell_type)) {
  test_mk1_mk09_5cells(celltype)
}

```

Get a table with the genes with at least 1 position with m/k = 1 in a cell type, and an m6a with m/k < 0.9 in at least one cell of another cell type

```{r echo = FALSE, message = FALSE, warning = FALSE}
test_mk1_mk09_genes <- function(celltype) {
  mk1_genes <- yth_data_cluster_labeled %>%
    filter(cell_type == celltype,
           dart_freq == 1) %>%
    select(gene) %>%
    distinct()

  common_genes <- yth_data_cluster_labeled %>%
    filter(cell_type != celltype) %>%
    semi_join(mk1_genes, by = c("gene")) %>%
    select(gene) %>%
    distinct()

  other_celltypes <- yth_data_cluster_labeled %>%
    filter(cell_type != celltype) %>%
    filter(dart_freq < 0.9)

  genes_with_other_mk09 <- mk1_genes %>%
    inner_join(other_celltypes, by = c("gene")) %>%
    group_by(gene) %>%
    summarise(n_cells = n_distinct(cell_id))


print(sprintf("%s genes have at least 1 position with m/k = 1 in %s", nrow(mk1_genes), celltype))
print(sprintf("Of those genes, %s are also found in at least one cell of another cell type", nrow(common_genes)))
print(sprintf("Of those %s genes, %s have an m6a with m/k < 0.9 in at least 1 cell of another cell type", nrow(distinct(common_genes)), nrow(filter(genes_with_other_mk09, n_cells > 0))))

# save the table per cell type
outdir <- dirname(filename)
write.table(genes_with_other_mk09, file = file.path(outdir, paste0(celltype, "_genes_mk1_common_mk09.txt")), sep = "\t", row.names = FALSE, quote = FALSE, col.names = TRUE)

}

# run test_mk1_mk09_genes for all cell types
for (celltype in unique(cluster_df$cell_type)) {
  test_mk1_mk09_genes(celltype)
}
```