---
title: "HEK_YTH-E_vs_E-YTH"
output:
  html_document:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script we compare C-to-U editing sites for YTH-E and E-YTH in HEK293T cell lines. 
For obtaining these variants we have:

In each replicate:

- Kept only variants with FDR<1, m>= 2 (two editing events), k >= 10 (coverage >= 10)
- Kept only variants with m/k >5%
- Discarded SNPs: dbSNP

In each sample (that consists of 3 replicates):

- Kept only those variants present in at least 2/3 replicates

Finally, we have:

- Removed from the YTH-mut sample the variants in APOBEC and WT
- Removed from YTH the variants in APOBEC and WT
- Kept in the YTH sample only those variants with a frequency at least 1.5X higher than in YTH-mut sample

### Distribution of m/k values:

```{r V1, echo = FALSE, warning=FALSE, message = FALSE, dev=c('png', 'pdf')}
library(tidyverse)
library(RColorBrewer)
dir <- "/home/Magda_lab/maitenat/scratch60/dartseq_replicates/aligned"

# Read data ----

yth_v1 <- read.table(file.path(dir, "HEK293T-V1_yth_no-back-apo.cims"), header = FALSE) %>%
  as_tibble()
colnames(yth_v1) <- c("chr", "start", "end", "name", "score", "strand", "m/k")

ythmut_v1 <- read.table(file.path(dir, "HEK293T-V1_ythmut_no-back-apo.cims"), header = FALSE) %>%
  as_tibble()
colnames(ythmut_v1) <- c("chr", "start", "end", "name", "score", "strand", "m/k")

yth_diff_v1 <- read.table(file.path(dir, "HEK293T-V1_yth_filt-all.cims"), header = FALSE, fill = TRUE) %>%
  as_tibble() %>%
  select(1:7)
colnames(yth_diff_v1) <- c("chr", "start", "end", "name", "score", "strand", "m/k")

```


```{r V2, echo = FALSE, warning=FALSE, message = FALSE, dev=c('png', 'pdf')}
yth_v2 <- read.table(file.path(dir, "HEK293T-V2_yth_no-back-apo.cims"), header = FALSE) %>%
  as_tibble()
colnames(yth_v2) <- c("chr", "start", "end", "name", "score", "strand", "m/k")

ythmut_v2 <- read.table(file.path(dir, "HEK293T-V2_ythmut_no-back-apo.cims"), header = FALSE) %>%
  as_tibble()
colnames(ythmut_v2) <- c("chr", "start", "end", "name", "score", "strand", "m/k")

yth_diff_v2 <- read.table(file.path(dir, "HEK293T-V2_yth_filt-all.cims"), header = FALSE, fill = TRUE) %>%
  as_tibble() %>%
  select(1:7)
colnames(yth_diff_v2) <- c("chr", "start", "end", "name", "score", "strand", "m/k")
```

```{r, echo = FALSE, warning=FALSE, message = FALSE, dev=c('png', 'pdf')}
all_diff <- bind_rows(v1 = yth_diff_v1, v2 = yth_diff_v2, .id = "id")

all_diff %>%
  ggplot(aes(`m/k`, fill = id, colour = id)) +
  geom_histogram(position = "identity", alpha=0.7, binwidth = 0.01) +
  scale_colour_manual(values = c("#9999CC", "#FFCCCC")) +
  scale_fill_manual(values = c("#9999CC", "#FFCCCC")) +
  theme_classic() 

```

### CDS plots

#### YTH-E

```{r, echo = FALSE, warning=FALSE, message = FALSE, dev=c('png', 'pdf')}
cds_file <- "/GPFS/Magda_lab_temp/maitenat/dartseq_replicates/aligned/metaplotR/HEK293T-V1_yth_filt-all.dist.measures.txt"
m6a.dist <- read.delim(cds_file, header = T)
trx_len <- m6a.dist$utr5_size + m6a.dist$cds_size + m6a.dist$utr3_size # Determine transcript length
temp <- data.frame(m6a.dist$gene_name, m6a.dist$refseqID, trx_len)
colnames(temp) <- c("gene_name", "gid", "trx_len") 
temp.df <- temp[order(temp$gene_name, temp$gid, -temp$trx_len),]
temp.df <- temp[!duplicated(temp$gene_name),]

# limit m6a data to one transcript per gene (longest)
m6a.dist <- m6a.dist[m6a.dist$refseqID %in% temp.df$gid,]

ggplot(m6a.dist, aes(x = rel_location)) +
  geom_histogram(fill="#9999CC") +
  theme_classic()

m6a.dist %>%
  ggplot(aes(x = utr3_st)) +
  geom_density(aes(y = ..count..), alpha=0.6, fill = "#8DA0CB", color = "#8DA0CB") +
  xlim (-500,500) +
  theme_classic()

# count the number of variants with rel_location between 0 and 1
sum(m6a.dist$rel_location >= 0 & m6a.dist$rel_location < 1)/nrow(m6a.dist)
# count the number of variants with rel_location between 1 and 2
sum(m6a.dist$rel_location >= 1 & m6a.dist$rel_location < 2)/nrow(m6a.dist)
# count the number of variants with rel_location between 2 and 3
sum(m6a.dist$rel_location >= 2 & m6a.dist$rel_location <= 3)/nrow(m6a.dist)
# percentage of variants in the first 10% of the 3UTR
sum(m6a.dist$rel_location >= 2 & m6a.dist$rel_location < 2.1)/nrow(m6a.dist)
```

#### E-YTH

```{r, echo = FALSE, warning=FALSE, message = FALSE, dev=c('png', 'pdf')}
cds_file <- "/GPFS/Magda_lab_temp/maitenat/dartseq_replicates/aligned/metaplotR/HEK293T-V2_yth_filt-all.dist.measures.txt"
m6a.dist <- read.delim(cds_file, header = T)
trx_len <- m6a.dist$utr5_size + m6a.dist$cds_size + m6a.dist$utr3_size # Determine transcript length
temp <- data.frame(m6a.dist$gene_name, m6a.dist$refseqID, trx_len)
colnames(temp) <- c("gene_name", "gid", "trx_len") 
temp.df <- temp[order(temp$gene_name, temp$gid, -temp$trx_len),]
temp.df <- temp[!duplicated(temp$gene_name),]

# limit m6a data to one transcript per gene (longest)
m6a.dist <- m6a.dist[m6a.dist$refseqID %in% temp.df$gid,]

ggplot(m6a.dist, aes(x = rel_location)) +
  geom_histogram(fill="#FFCCCC") +
  theme_classic()

m6a.dist %>%
  ggplot(aes(x = utr3_st)) +
  geom_density(aes(y = ..count..), alpha=0.6, fill = "#8DA0CB", color = "#8DA0CB") +
  xlim (-500,500) +
  theme_classic()

# count the number of variants with rel_location between 0 and 1
sum(m6a.dist$rel_location >= 0 & m6a.dist$rel_location < 1)/nrow(m6a.dist)
# count the number of variants with rel_location between 1 and 2
sum(m6a.dist$rel_location >= 1 & m6a.dist$rel_location < 2)/nrow(m6a.dist)
# count the number of variants with rel_location between 2 and 3
sum(m6a.dist$rel_location >= 2 & m6a.dist$rel_location <= 3)/nrow(m6a.dist)
# percentage of variants in the first 10% of the 3UTR
sum(m6a.dist$rel_location >= 2 & m6a.dist$rel_location < 2.1)/nrow(m6a.dist)
```

### Localization plots

#### YTH-E

```{r, echo = FALSE, warning=FALSE, message = FALSE}
blank_theme <- theme_minimal() +
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

yth_annot <- read.delim(file.path(dir, "HEK293T-V1_yth_filt-all_annot.bed"), header = TRUE, sep = "\t") %>%
  as_tibble()

yth_annot <- yth_annot %>%
  mutate(annot = gsub("\\(.*", "", Annotation))

yth_annot_table <- table(yth_annot$annot)
yth_annot_table0 <- tibble(localization = names(yth_annot_table), variants = as.numeric(yth_annot_table))
  
ggplot(yth_annot_table0, aes(x = "", y = variants, fill = localization)) +
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) +
  blank_theme +
  theme(axis.text.x = element_blank())

yth_annot_table0 %>%
  mutate(variant_percentage = variants/sum(variants))
```

#### E-YTH

```{r, echo = FALSE, warning=FALSE, message = FALSE}
blank_theme <- theme_minimal() +
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

yth_annot <- read.delim(file.path(dir, "HEK293T-V2_yth_filt-all_default_annot.bed"), header = TRUE, sep = "\t") %>%
  as_tibble()

yth_annot <- yth_annot %>%
  mutate(annot = gsub("\\(.*", "", Annotation))

yth_annot_table <- table(yth_annot$annot)
yth_annot_table0 <- tibble(localization = names(yth_annot_table), variants = as.numeric(yth_annot_table))
  
ggplot(yth_annot_table0, aes(x = "", y = variants, fill = localization)) +
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) +
  blank_theme +
  theme(axis.text.x = element_blank())

yth_annot_table0 %>%
  mutate(variant_percentage = variants/sum(variants))

```